\subsection{Probability models and axioms}

A probabilistic model describes the behavior of a random experiment, such as flipping a coin, rolling dice, or throwing a dart. Building such a model has two steps: first, we list all possible outcomes; second, we assign probabilities to them. In this section, we focus on the first step — describing the possible outcomes clearly and systematically.

\subsubsection{Sample space}

All possible outcomes of an experiment form a set, denoted by \(\Omega\), called the sample space. Each element of \(\Omega\) represents one possible result. The outcomes must be mutually exclusive (only one can happen at a time) and collectively exhaustive (some outcome from \(\Omega\) must always happen).

How detailed our sample space is depends on what we care about in the experiment. It should include all relevant distinctions but ignore unimportant ones. Choosing a good sample space is part of the modeling process.

\bigskip
If we flip a coin, a simple sample space is
\[
\Omega = \{\text{heads}, \text{tails}\}.
\]
This covers all outcomes and they don’t overlap.

If we also record whether it’s raining, we could define a larger sample space,
\[
\Omega' = \{\text{heads and rain}, \text{heads and no rain}, \text{tails and rain}, \text{tails and no rain}\}.
\]
This is valid too, but since weather usually doesn’t affect a coin toss, it’s an unnecessary detail. We include such extra factors only when they matter for the problem.

\subsubsection{Discrete and continuous sample spaces}

Sample spaces can be discrete or continuous.

For example, rolling a four-sided die twice and recording the two results gives
\[
\Omega = \{(i,j): i,j \in \{1,2,3,4\}\},
\]
with 16 possible ordered pairs like \((1,1)\) or \((3,4)\). When experiments have several stages, tree diagrams can help visualize the structure of outcomes.

Other sample spaces are continuous. For instance, when throwing a dart at a unit square, each landing point is a pair of coordinates \((x,y)\):
\[
\Omega = \{(x,y): 0 \le x \le 1,\ 0 \le y \le 1\}.
\]
Here, every point inside the square is a possible outcome.

\subsubsection{Events and probability laws}

Once we have a sample space, we can talk about events. An event \(A\) is any subset of \(\Omega\). After an experiment, if the observed outcome \(\omega\) belongs to \(A\), we say the event occurred.

In discrete cases, we can assign probabilities to single outcomes. In continuous cases, each exact point has probability 0, so we assign probabilities to regions instead (for example, “the dart lands in the upper half of the square”). The probability law gives each event \(A\) a value \(P(A)\), showing how likely that event is.

\subsubsection{Probability axioms and basic set notation}

Probabilities follow three basic rules, called axioms:

\begin{itemize}
    \item \textbf{Nonnegativity}: For any event \(A\), \(P(A) \ge 0\).
    \item \textbf{Normalization}: \(P(\Omega) = 1\), since something in the sample space must occur.
    \item \textbf{Additivity}: If two events \(A\) and \(B\) are disjoint (that is, \(A \cap B = \emptyset\)), then
    \[
    P(A \cup B) = P(A) + P(B).
    \]
\end{itemize}

You can think of probability as a kind of “mass” spread over \(\Omega\). If two events don’t overlap, the total mass on their union is just the sum of their individual masses.

\newpage
For two events \(A\) and \(B\):
\begin{itemize}
    \item The intersection \(A \cap B\) is the set of outcomes that belong to both \(A\) and \(B\).
    \item The union \(A \cup B\) is the set of outcomes that belong to either \(A\) or \(B\) (or both).
    \item Two events are disjoint if \(A \cap B = \emptyset\), meaning they have no outcomes in common.
\end{itemize}


\subsubsection{Simple properties derived from the axioms}

The three axioms look simple, but they already imply many useful and intuitive properties. For example, we can prove that probabilities are always at most 1, and that the probability of the empty set is 0, instead of assuming these as extra axioms.

Consider an event \(A\) and its complement \(A^c\), which contains all outcomes in \(\Omega\) that are not in \(A\). The set \(A\) and its complement \(A^c\) are disjoint, and their union is the entire sample space:
\[
A \cup A^c = \Omega, \qquad A \cap A^c = \emptyset.
\]
Using normalization and additivity, we get
\[
P(\Omega) = P(A \cup A^c) = P(A) + P(A^c) = 1.
\]
From this relation we also obtain
\[
P(A) = 1 - P(A^c).
\]
Because probabilities are nonnegative, \(P(A^c) \ge 0\), so \(P(A) \le 1\). This shows that probabilities are always bounded above by 1, without adding a separate axiom.

If we apply the same relation to the special case where \(A = \Omega\), then the complement \(A^c\) is the empty set \(\emptyset\). We already know that \(P(\Omega) = 1\), so
\[
1 = P(\Omega) = P(\Omega) + P(\emptyset),
\]
which forces \(P(\emptyset) = 0\). This matches the idea that the empty set is “impossible,” so its probability should be zero.

Finally, the additivity axiom for two disjoint events extends to any finite collection of disjoint events. If \(A_1, A_2, \dots, A_k\) are pairwise disjoint, then
\[
P\Big(\bigcup_{i=1}^k A_i\Big) = \sum_{i=1}^k P(A_i).
\]
A common special case is when we have a finite subset of the sample space consisting of points \(s_1, \dots, s_k\). We can view this set as the union of the singletons \(\{s_1\}, \dots, \{s_k\}\), which are disjoint. Then the probability of the finite set is just the sum of the probabilities of its individual elements. In practice we often write \(P(s_i)\) instead of \(P(\{s_i\})\) and talk informally about “the probability of an outcome \(s_i\)” even though probabilities are formally defined on sets.

\bigskip

The axioms also give us some useful monotonicity and union properties. Suppose we have two events \(A\) and \(B\) with \(A \subseteq B\). Intuitively, \(B\) is “larger,” so its probability should be at least as big as that of \(A\). Formally, we can write
\[
B = A \cup \big(B \cap A^c\big),
\]
where \(A\) and \(B \cap A^c\) are disjoint. By additivity,
\[
P(B) = P(A) + P\big(B \cap A^c\big) \ge P(A),
\]
since probabilities are nonnegative.

We can also express the probability of a union when the sets are not disjoint. If \(A\) and \(B\) overlap, then the union \(A \cup B\) can be decomposed into three disjoint parts: the part of \(A\) outside \(B\), the intersection \(A \cap B\), and the part of \(B\) outside \(A\). If we call their probabilities \(a\), \(b\), and \(c\), then
\[
P(A \cup B) = a + b + c.
\]
On the other hand,
\[
P(A) = a + b, \qquad P(B) = b + c, \qquad P(A \cap B) = b.
\]
Putting these together gives the standard identity
\[
P(A \cup B) = P(A) + P(B) - P(A \cap B).
\]

A direct corollary is the union bound. Since \(P(A \cap B) \ge 0\), we get
\[
P(A \cup B) \le P(A) + P(B),
\]
which is useful when we only need an upper bound on a probability.

The same idea extends to three sets. For events \(A\), \(B\), and \(C\), we can rewrite their union as a union of three disjoint pieces: \(A\) itself, the part of \(B\) outside \(A\), and the part of \(C\) outside both \(A\) and \(B\). Using additivity on these disjoint parts gives a formula for \(P(A \cup B \cup C)\) in terms of simpler probabilities. More generally, we can keep breaking unions into disjoint pieces to get expressions for probabilities of unions of several events.

\subsubsection{A discrete example and the discrete uniform law}

Consider again two rolls of a tetrahedral die. The sample space has 16 outcomes, each an ordered pair \((X,Y)\), where \(X\) is the first roll and \(Y\) is the second. Assume all 16 outcomes are equally likely, so each has probability \(1/16\).

As a first example, look at the event \(\{X = 1\}\): the first roll is 1. In the grid of all \((X,Y)\) pairs, this event contains the four outcomes where the first coordinate is 1 and the second is 1, 2, 3, or 4. Since each outcome has probability \(1/16\), we get
\[
P(X = 1) = 4 \cdot \frac{1}{16} = \frac{1}{4}.
\]

Now define \(Z\) to be the smaller of the two numbers in the two rolls. For example, if \(X = 2\) and \(Y = 3\), then \(Z = 2\). For \(P(Z = 4)\), this happens only if both \(X\) and \(Y\) are 4, so there is exactly one outcome in the event and
\[
P(Z = 4) = \frac{1}{16}.
\]
For \(\{Z = 2\}\), we need one roll equal to 2 and the other at least 2. This event contains \((2,2)\), \((2,3)\), \((2,4)\), \((3,2)\), and \((4,2)\), for a total of five outcomes. Thus
\[
P(Z = 2) = 5 \cdot \frac{1}{16}.
\]

This is an instance of the discrete uniform law. If the sample space \(\Omega\) is finite with \(n\) elements and all are equally likely, then, since \(P(\Omega) = 1\), we must have
\[
P(\{\omega\}) = \frac{1}{n}
\]
for each outcome \(\omega\). If an event \(A\) has exactly \(k\) elements, then
\[
P(A) = k \cdot \frac{1}{n} = \frac{k}{n}.
\]
So in a discrete uniform model, computing probabilities amounts to counting: we count all possible outcomes \(n\), and the favorable ones \(k\), then take the ratio \(k/n\).

\subsubsection{A continuous example}

Now consider a probability calculation in a continuous sample space. The experiment is again throwing a dart at the unit square, and the outcome is the point \((x,y)\) where the dart lands. The sample space is the unit square, and we assume a uniform law: the probability of any region in the square is equal to its area. This is natural if we think the dart is equally likely to land anywhere.

Take the event that the sum of the coordinates is at most \(1/2\):
\[
A = \{(x,y) : x + y \le 1/2\}.
\]
In the unit square this is a right triangle under the line \(x + y = 1/2\). Since probability equals area,
\[
P(A) = \frac{1}{2} \cdot \frac{1}{2} \cdot \frac{1}{2} = \frac{1}{8}.
\]
As another example, consider the event that the dart hits exactly the point \((0.5, 0.3)\). This event is a single point with area 0, so its probability is 0. The same holds for any single point in the unit square: in a continuous model, individual points have probability 0, while regions with positive area can have positive probability.

This example fits a general four-step recipe for probability problems. Starting from a word description of an experiment, we:
\begin{enumerate}
    \item Specify the sample space.
    \item Choose a probability law (a model for how likely different outcomes are).
    \item Describe the event of interest precisely, ideally also with a picture.
    \item Use the probability law and the event description to compute the desired probability.
\end{enumerate}
Sometimes the law is given explicitly for all events; in other cases we only know it for some basic events and need extra work to obtain the probability we want.

\subsubsection{Countable additivity and infinite sample spaces}

We now look at a discrete but infinite sample space. Imagine an experiment whose outcome is a positive integer, for example the number of coin tosses until the first heads appears. The sample space is \(\{1,2,3,\dots\}\). Suppose we are told that
\[
P(\{n\}) = \frac{1}{2^n}, \quad n = 1,2,3,\dots
\]
First we check that this is a valid probability law by summing over all \(n\):
\[
\sum_{n=1}^\infty \frac{1}{2^n}
= \frac{1}{2} \sum_{n=0}^\infty \Big(\frac{1}{2}\Big)^n
= \frac{1}{2} \cdot \frac{1}{1 - 1/2} = 1.
\]

Now consider the event that the outcome is even, the infinite set \(\{2,4,6,\dots\}\). We can write it as a union of singletons:
\[
\{2,4,6,\dots\} = \{2\} \cup \{4\} \cup \{6\} \cup \dots
\]
These sets are disjoint, so we would like to write
\[
P(\text{even}) = P(\{2\}) + P(\{4\}) + P(\{6\}) + \dots
\]
and evaluate the resulting infinite series. To make this step valid in general, we extend the additivity axiom from finite unions to countably infinite unions.

The extra axiom, called countable additivity, says: if \(A_1,A_2,A_3,\dots\) is a sequence of pairwise disjoint events, then
\[
P\Big(\bigcup_{n=1}^\infty A_n\Big) = \sum_{n=1}^\infty P(A_n).
\]
The word “sequence” matters: the events can be listed in order \(A_1,A_2,\dots\), so we can form a countable sum. This axiom lets us compute probabilities of many events in infinite sample spaces by summing over simpler disjoint pieces.

It is also important to see what countable additivity does not say. In the continuous unit square model, the square is the union of singletons (one per point), and each singleton has probability 0. If we tried to add their probabilities, we would get a sum of zeros, but the probability of the whole square should be 1. The catch is that the unit square is uncountable: its points cannot be arranged as a sequence \(A_1,A_2,\dots\), so the axiom does not apply. In short, some infinite sets are countable (like the integers) and others are uncountable (like the real line or the unit square), and countable additivity only covers the countable case.

There are deeper measure-theoretic issues in the background, especially when we try to define probability on all subsets of a continuous space. For this course, the key point is that standard constructions, like defining probability on the unit square via area, can be made rigorous and do satisfy countable additivity for all “nice” sets we will use. The strange sets where problems arise are mostly of theoretical interest and do not appear in typical applications.

\subsubsection{Interpretations and uses of probability}

To close, it is useful to think about what probability really means. In the narrow sense, probability theory is a branch of mathematics: we start from axioms, study models that satisfy them, and prove theorems. This view does not fix any particular real-world interpretation.

One common interpretation is in terms of long-run frequencies. For example, if we have a fair coin and toss it many times, the fraction of heads is expected to approach \(1/2\). In this view, the probability of an event is the limiting frequency with which it occurs in a long sequence of repeated trials. This fits well for repeatable experiments like coin tosses.

But not all probability statements are like that. A sentence such as “the current president of my country will be reelected with probability \(0.7\)” does not describe an experiment we can repeat infinitely many times. In such cases it is more natural to see probabilities as measuring degrees of belief: they summarize how strongly we believe an event will happen. Thinking in terms of bets, a probability expresses how much we are willing to stake on an outcome.

This raises the worry that probabilities are just subjective. The answer is that probability theory at least gives consistent rules for reasoning under uncertainty. If our probability model is a reasonable description of the situation, those rules help us make predictions and decisions in a systematic way.

Whether our decisions are actually good depends on whether we chose a good model. Building such models from data is the task of statistics and statistical inference. The real world generates data; statistics uses these data to construct probabilistic models; and probability theory then analyzes these models to produce predictions or decisions that we can apply back to the real world.


\subsubsection{Sets, complement, empty set, and subsets}

A set is a collection of distinct elements, usually denoted by a capital letter like \(S\). We can specify a set by listing its elements inside braces, for example \(\{a,b,c,d\}\), which is a finite set. Other sets, such as the set of all real numbers, are infinite. We write \(x \in S\) if \(x\) is an element of \(S\), and \(x \notin S\) if \(x\) does not belong to \(S\).

Sometimes it is convenient to define a set by a property instead of listing elements. For example, starting from the real numbers, we can consider all \(x\) such that \(\cos(x) > 1/2\); the collection of all such \(x\) is a set. Often we also fix a larger “universe” of objects, called the universal set and denoted by \(\Omega\), and only look at sets contained in this universe.

Given a universal set \(\Omega\) and a subset \(S \subseteq \Omega\), the complement of \(S\), denoted \(S^c\), is the set of all elements in \(\Omega\) that are not in \(S\):
\[
S^c = \{x \in \Omega : x \notin S\}.
\]
Taking complements twice brings us back to the original set: \((S^c)^c = S\). A special set is the empty set \(\emptyset\), which contains no elements. For example, the complement of \(\Omega\) is \(\emptyset\), since there is nothing outside the universal set.

If every element of a set \(S\) also belongs to a set \(T\), we say that \(S\) is a subset of \(T\) and write \(S \subseteq T\). This includes the case \(S = T\). Intuitively, \(S\) is “contained” inside \(T\).

\subsubsection{Union, intersection, and infinite collections}

Given two sets \(S\) and \(T\), their union \(S \cup T\) is the set of all elements that belong to at least one of them:
\[
x \in S \cup T \quad \text{iff} \quad x \in S \ \text{or} \ x \in T.
\]
Their intersection \(S \cap T\) is the set of elements that belong to both:
\[
x \in S \cap T \quad \text{iff} \quad x \in S \ \text{and} \ x \in T.
\]

These notions extend to infinite families of sets. If we have sets \(S_1, S_2, S_3, \dots\), their union is
\[
\bigcup_{n=1}^\infty S_n = \{x : x \in S_n \ \text{for at least one } n\},
\]
and their intersection is
\[
\bigcap_{n=1}^\infty S_n = \{x : x \in S_n \ \text{for all } n\}.
\]
Informally, the union collects elements that appear in at least one of the sets, while the intersection keeps only the elements common to all of them.

\subsubsection{Basic algebra of sets}

Set operations satisfy some familiar algebraic properties. Union and intersection are both commutative:
\[
S \cup T = T \cup S, \qquad S \cap T = T \cap S,
\]
and associative:
\[
(S \cup T) \cup U = S \cup (T \cup U), \qquad (S \cap T) \cap U = S \cap (T \cap U).
\]
Because of associativity, we usually drop parentheses when writing unions or intersections of three or more sets.

The universal set \(\Omega\) behaves like an identity: \(S \cup \Omega = \Omega\) and \(S \cap \Omega = S\). There are also distributive laws describing how union and intersection interact, for example
\[
S \cap (T \cup U) = (S \cap T) \cup (S \cap U),
\]
and a similar law with \(\cup\) and \(\cap\) swapped. A standard way to check such identities is to argue element-wise: assume \(x\) belongs to one side, unpack the definitions, and show that this forces \(x\) to belong to the other side as well. If each side is a subset of the other, then the two sets are equal.

\subsubsection{De Morgan's laws}

De Morgan's laws relate unions, intersections, and complements. For two sets \(S\) and \(T\), the first law says that the complement of their intersection is the union of their complements:
\[
(S \cap T)^c = S^c \cup T^c.
\]
Intuitively, if a point is not in both \(S\) and \(T\) at the same time, then it must be missing from at least one of them, so it lies in \(S^c\) or in \(T^c\). A formal proof rewrites the membership condition step by step: \(x \notin S \cap T\) means “\(x\) is not in \(S\) or not in \(T\),” which is the same as “\(x \in S^c\) or \(x \in T^c\),” that is, \(x \in S^c \cup T^c\).

The second law is symmetric and says that the complement of a union is the intersection of the complements:
\[
(S \cup T)^c = S^c \cap T^c.
\]
Here the idea is that if a point is not in \(S \cup T\), then it cannot be in \(S\) and cannot be in \(T\), so it must belong to both complements. Both identities extend to families of sets \(\{S_n\}\):
\[
\Big(\bigcap_{n} S_n\Big)^c = \bigcup_{n} S_n^c,
\qquad
\Big(\bigcup_{n} S_n\Big)^c = \bigcap_{n} S_n^c.
\]
These laws are very useful when we want to switch between unions and intersections, especially in probability.
\subsubsection{Sequences and their limits}

A sequence is a collection of elements indexed by the positive integers. We usually write it as \(\{a_i\}_{i=1}^\infty\), or simply \(a_i\), where each \(a_i\) belongs to some set \(S\). Often \(S = {R}\) and we have a sequence of real numbers, but \(S\) could also be \({R}^n\) (a sequence of vectors) or some other set. Formally, a sequence is just a function \(f\) from the natural numbers to \(S\), and the \(i\)-th term is \(a_i = f(i)\).

We say that a sequence \(a_i\) converges to a real number \(a\) if, as \(i\) becomes large, the terms of the sequence get arbitrarily close to \(a\) and then stay close. Formally, \(a_i \to a\) as \(i \to \infty\) if for every \(\varepsilon > 0\) there exists an index \(i_0\) such that for all \(i \ge i_0\),
\[
|a_i - a| < \varepsilon.
\]
Geometrically, if you draw a horizontal band of width \(2\varepsilon\) around the limit \(a\), then from some point on all the sequence terms lie inside that band. The definition may look technical, but it just encodes the idea that the sequence eventually stays as close to \(a\) as we like.

Convergence behaves well under basic operations. If \(a_i \to a\) and \(b_i \to b\), then the sum sequence \(a_i + b_i\) converges to \(a + b\), and the product sequence \(a_i b_i\) converges to \(ab\). If \(g\) is a continuous function and \(a_i \to a\), then the sequence \(g(a_i)\) converges to \(g(a)\); for example, if \(a_i \to a\), then \(a_i^2 \to a^2\). These properties let us manipulate limits in a straightforward way.

\subsubsection{Practical convergence criteria}

In practice, it is helpful to have quick tests for convergence. One common situation is a monotone sequence. If \(a_i\) is increasing (or at least never decreases) and is bounded above, then it converges to some finite limit. If it keeps increasing without any bound, we say that it “converges to infinity” rather than to a finite number.

Another strategy is to bound the distance \(|a_i - a|\) from a candidate limit \(a\). If we can find an upper bound on \(|a_i - a|\) that goes to 0 as \(i\) grows, then \(a_i\) must converge to \(a\). A related idea is the sandwich (or squeeze) argument: if a sequence \(a_i\) is always between two other sequences \(b_i\) and \(c_i\), and both \(b_i\) and \(c_i\) converge to the same limit \(a\), then \(a_i\) also converges to \(a\). These tools will appear often, sometimes implicitly, when analyzing limits in the course.

\subsubsection{Infinite Series}

Consider a sequence \(a_i\) and the sum:
\[
\sum_{i=1}^{\infty} a_i.
\]
This sum is defined as the limit of the partial sums:
\[
S_n = \sum_{i=1}^n a_i,
\]
if the limit exists and is finite as \(n \to \infty\). The series converges if this limit exists; otherwise, it diverges.

For non-negative terms, partial sums are increasing and bounded, ensuring convergence (finite or infinite). For alternating signs, the limit may depend on the order of terms, so absolute convergence is preferred. If
\[
\sum_{i=1}^{\infty} |a_i|
\]
converges, then the series \(\sum a_i\) converges as well.

A key example is the geometric series. For \(|\alpha| < 1\), consider:
\[
\sum_{i=0}^{\infty} \alpha^i = 1 + \alpha + \alpha^2 + \dots
\]
The sum is:
\[
\sum_{i=0}^{\infty} \alpha^i = \frac{1}{1 - \alpha}.
\]
Alternatively, using the trick \(s = \sum_{i=0}^{\infty} \alpha^i\), we get:
\[
s = \frac{1}{1 - \alpha}, \quad |\alpha| < 1.
\]


\subsubsection{Series with Multiple Indices and Order of Summation}

For series with two indices, like \(a_{ij}\) with \(i,j = 1,2,\dots\), we sum over a grid of points \((i,j)\). The sum can be written as:
\[
\sum_{i=1}^{\infty} \sum_{j=1}^{\infty} a_{ij} \quad \text{or} \quad \sum_{j=1}^{\infty} \sum_{i=1}^{\infty} a_{ij}.
\]
Different summation orders may yield different results, but absolute convergence ensures the sum is well-defined, and the order does not matter:
\[
\sum_{i=1}^{\infty} \sum_{j=1}^{\infty} |a_{ij}|.
\]
Without absolute convergence, changing the order can lead to different sums, such as 0 or 1, due to cancellations between positive and negative terms.

We can also sum over restricted sets, e.g., \(\sum_{i=1}^{\infty} \sum_{j=1}^{i} a_{ij}\), and the order of summation can vary. With absolute convergence, the result is independent of the summation order.

\subsubsection{Countable and Uncountable Sets}

In probability theory, not all infinite sets are the same. Some are countable, others uncountable. A set is countable if its elements can be mapped one-to-one to natural numbers, i.e., they can be listed in a sequence.

Positive integers are countable, as are all integers, which can be listed as \(0, 1, -1, 2, -2, 3, -3, \dots\). The same applies to pairs \((i,j)\) of positive integers, which can be arranged in a zigzag pattern.

A less obvious example is the set of rationals in \([0,1]\). Since every rational can be written as \(p/q\), we can list them by varying \(q\) and \(p\), skipping duplicates, resulting in a sequence containing all rationals in \([0,1]\).

A set is uncountable if it cannot be listed in this way. Typical examples include intervals of real numbers, the real line, and spaces like the plane or 3D space. These sets are too large to be listed. The distinction between countable and uncountable sets is key to understanding the difference between discrete and continuous probability models.


\subsubsection{Cantor’s Diagonal Argument}

Cantor's diagonal argument proves that the set of real numbers in \([0,1]\) is uncountable. The idea is to assume that we can list all real numbers in this interval, each represented by an infinite decimal expansion. If we arrange them in a sequence, like:

\[
r_1 = 0.a_{11}a_{12}a_{13}\dots, \quad r_2 = 0.a_{21}a_{22}a_{23}\dots, \quad r_3 = 0.a_{31}a_{32}a_{33}\dots, \dots
\]

Then, by constructing a new number \(r\) whose \(n\)-th decimal place is different from the \(n\)-th decimal of \(r_n\), we ensure that \(r\) is not on the list. This contradiction shows that the real numbers in \([0,1]\) cannot be counted, proving they are uncountable.

Thus, Cantor's argument demonstrates that no list can exhaust all real numbers, highlighting the distinction between countable and uncountable sets.