\subsection{Continuous random variables}

A continuous random variable takes values over a continuous range, such as the real line or an interval on the real line.  
In the discrete case, one unit of probability is distributed over the possible values as point masses, and the PMF tells us how much probability sits on top of each value.  
In the continuous case, one unit of probability is still assigned to the possible values, but now it is spread continuously over the real line.

Some parts of the real line receive more probability mass per unit length, others receive less.  
The \textit{probability density function} (PDF), denoted \(f_X(x)\), describes how densely probability is distributed along the real line.  
The letter \(f\) indicates we are dealing with a PDF, and the subscript indicates which random variable we are considering.

\bigskip\textsc{Interpretation of the PDF}

Think of probability as snowfall.  
One pound of snow has fallen on the real line, and the PDF tells us the height of the snow accumulated over each point.  
The weight of snow sitting on top of an interval \([a, b]\) is found by calculating the area under the PDF curve over that interval.  
Mathematically, area under the curve is an integral, so
\[
\mathbb{P}(a \le X \le b) = \int_a^b f_X(x) \, dx.
\]
This is the probability that \(X\) takes values in the interval \([a, b]\).

\bigskip\textsc{Properties of a PDF}

By analogy with the discrete case, a PDF must satisfy two conditions:
\[
f_X(x) \ge 0 \quad \text{for all } x,
\]
\[
\int_{-\infty}^{+\infty} f_X(x) \, dx = 1.
\]
The first condition ensures we do not obtain negative probabilities.  
The second condition ensures that the total probability over the entire real line equals one, since \(X\) is certain to lie somewhere on the real line.  
These two conditions are all that we need for a legitimate PDF.

\bigskip\textsc{Formal definition of a continuous random variable}

A continuous random variable is a random variable whose probabilities can be described by a PDF.  
That is, there exists a function \(f_X\) such that
\[
\mathbb{P}(a \le X \le b) = \int_a^b f_X(x) \, dx
\]
for any interval \([a, b]\).  
It is not enough for a random variable to take values on a continuous set.  
A continuous random variable must also be describable by a PDF according to this formula.

\bigskip\textsc{Probabilities of more complicated sets}

Once we have the probabilities of intervals, we can use additivity to calculate the probabilities of more complicated sets.  
For example, suppose we want the probability that \(X\) lies between \(1\) and \(3\) or between \(4\) and \(5\).  
This is the probability of a region consisting of two disjoint intervals.  
By additivity,
\[
\mathbb{P}(1 \le X \le 3 \text{ or } 4 \le X \le 5) = \mathbb{P}(1 \le X \le 3) + \mathbb{P}(4 \le X \le 5).
\]
Each of the probabilities on the right is calculated using the PDF:
\[
= \int_1^3 f_X(x) \, dx + \int_4^5 f_X(x) \, dx.
\]

\bigskip\textsc{The role of the sample space}

There is still an underlying sample space lurking in the background.  
Different outcomes in the sample space result in different numerical values for the random variable \(X\).  
When we talk about the probability that \(X\) takes values in an interval \([a, b]\), we mean the probability of those outcomes for which the resulting value of \(X\) lies in that interval.

However, once we have a PDF in our hands, we can completely forget about the sample space and carry out any calculations just by working with the PDF.  
As we move on in this course, the sample space will appear less and less, and we will work directly with PDFs or PMFs.

\newpage
\bigskip\textsc{Probabilities of small intervals}

Let us look at an interval that starts at some point \(a\) and goes up to \(a + \delta\), where \(\delta\) is a small positive number.  
The probability that \(X\) falls in this interval is the area under the PDF over the interval \([a, a + \delta]\).  
As long as \(f_X\) does not change too much over this little interval, which will be the case if \(f_X\) is continuous, we can approximate the area by the area of a rectangle with constant height.  
The area of this rectangle is
\[
\mathbb{P}(a \le X \le a + \delta) \approx f_X(a) \cdot \delta.
\]
This gives us an interpretation of PDFs in terms of probabilities of small intervals.

If we rearrange this approximation, we see that
\[
f_X(a) \approx \frac{\mathbb{P}(a \le X \le a + \delta)}{\delta}.
\]
So the value of the PDF can be interpreted as \textit{probability per unit length}.  
PDFs are not probabilities, they are densities, and their units are probability per unit length.

\bigskip\textsc{Probability of a single point}

If the probability per unit length is finite and the length \(\delta\) is sent to zero, we obtain zero probability.  
More formally, if we let \(b = a\) in the integral
\[
\mathbb{P}(a \le X \le a) = \int_a^a f_X(x) \, dx,
\]
we get an integral over a zero-length interval, and that integral is zero.  
Therefore,
\[
\mathbb{P}(X = a) = 0
\]
for any particular point \(a\).

So for a continuous random variable, any particular point has zero probability.  
Yet, collectively, the infinitely many points in an interval together have positive probability.  
This is not a puzzle, it is exactly what happens with the ordinary notion of length.  
Single points have zero length, yet by putting together lots of points we can create a set with positive length.

\bigskip\textsc{Open and closed intervals}

A final consequence of the fact that individual points have zero probability is that it does not matter whether we include the endpoints of an interval or not.  
Using additivity,
\[
\mathbb{P}(a \le X \le b) = \mathbb{P}(X = a) + \mathbb{P}(a < X < b) + \mathbb{P}(X = b).
\]
Since \(\mathbb{P}(X = a) = 0\) and \(\mathbb{P}(X = b) = 0\), we conclude that
\[
\mathbb{P}(a \le X \le b) = \mathbb{P}(a < X < b).
\]
When calculating probabilities of intervals, it does not matter whether we include the endpoints or not.


\subsubsection{Uniform and piecewise constant PDFs}

The uniform random variable is a basic example of a continuous random variable.  
In the discrete uniform case, the possible values are integers between \(a\) and \(b\), all equally likely.  
In the continuous case, any real number between \(a\) and \(b\) is possible, and the PDF has the same height at all points in this interval.

Intuitively, a uniform random variable models a situation where we know that the value lies between \(a\) and \(b\), but we have no reason to believe that any location inside the interval is more likely than another.  
In this sense, it represents a situation of complete ignorance.

\bigskip\textsc{PDF of a uniform random variable}

Let \(X\) be uniform on \([a,b]\).  
Its PDF is constant on \([a,b]\) and zero elsewhere:
\[
f_X(x) =
\begin{cases}
c, & a \le x \le b, \\
0, & \text{otherwise}.
\end{cases}
\]
The total probability must be \(1\), so the area under the PDF is
\[
1 = \int_a^b c \, dx = c(b-a),
\]
which gives \(c = \frac{1}{b-a}\).  
Thus
\[
f_X(x) =
\begin{cases}
\frac{1}{b-a}, & a \le x \le b, \\
0, & \text{otherwise}.
\end{cases}
\]
Any two intervals of the same length inside \([a,b]\) have the same probability, because their probabilities are the base (length) times the same constant height.

\bigskip\textsc{Piecewise constant PDFs}

A more general kind of PDF is piecewise constant.  
In this case, the PDF is constant on each of several subintervals, possibly with different heights on different intervals.  
This shows that PDFs need not be continuous; they may have jumps and discontinuities.

For such a PDF to be valid, the total area under the curve must be \(1\).  
This total area is the sum of the areas of the rectangles corresponding to each constant piece.

\bigskip\textsc{Computing probabilities}

With a piecewise constant PDF, probabilities are obtained by adding areas of rectangles.  
If a certain event corresponds to an interval that intersects several constant pieces, its probability is the sum of the areas over each part of that interval.  
For example, if over two subintervals the heights are \(h_1\) and \(h_2\) and the lengths are \(\ell_1\) and \(\ell_2\), then the probability of their union is
\[
h_1 \ell_1 + h_2 \ell_2.
\]


\subsubsection{Means and variances}

We now develop continuous counterparts of the concepts of expectation and variance.  
In the discrete case, expectation is a weighted average of the possible values of a random variable, weighted by their probabilities.  
In the continuous case, we use the same idea, but the sum is replaced by an integral and the weights come from the PDF.

\bigskip\textsc{Expectation of a continuous random variable}

Let \(X\) be a continuous random variable with PDF \(f_X(x)\).  
Its expectation (or mean) is defined by
\[
\mathbb{E}[X] = \int_{-\infty}^{+\infty} x f_X(x) \, dx.
\]
This is a weighted average of the possible values of \(X\), where points with higher density \(f_X(x)\) contribute more.  
More generally, for any function \(g\),
\[
\mathbb{E}[g(X)] = \int_{-\infty}^{+\infty} g(x) f_X(x) \, dx.
\]
This is the continuous analogue of the discrete expected value rule.  

To ensure that the expectation is well defined, we usually assume that
\[
\int_{-\infty}^{+\infty} |x| f_X(x) \, dx < \infty.
\]
Under this assumption, \(\mathbb{E}[X]\) exists.  

The intuition of expectation remains the same as in the discrete case.  
It represents the average value we would see if we repeated the experiment many times independently.  
It can also be viewed as the center of gravity of the distribution.  
If the PDF is symmetric around a point \(c\), then the expectation is \(c\).

\bigskip\textsc{Basic properties of expectation}

Expectations of continuous random variables satisfy the same properties as in the discrete case.  
If \(X \ge 0\) always, then \(\mathbb{E}[X] \ge 0\).  
If \(X\) always lies in an interval \([a,b]\), then \(\mathbb{E}[X]\) also lies in \([a,b]\).

A key property is linearity of expectation.  
For constants \(a\) and \(b\),
\[
\mathbb{E}[aX + b] = a\,\mathbb{E}[X] + b.
\]
Linearity holds regardless of whether \(X\) is discrete or continuous.  

\bigskip\textsc{Variance and standard deviation}

The variance of a continuous random variable is defined exactly as in the discrete case:
\[
\mathrm{Var}(X) = \mathbb{E}\big[(X - \mu)^2\big],
\]
where \(\mu = \mathbb{E}[X]\).  
Using the expected value rule, this becomes
\[
\mathrm{Var}(X) = \int_{-\infty}^{+\infty} (x - \mu)^2 f_X(x) \, dx.
\]
The standard deviation is
\[
\sigma = \sqrt{\mathrm{Var}(X)}.
\]

Variance measures how spread out the values of \(X\) are around the mean.  
A small variance means that most of the probability mass is concentrated close to \(\mu\).  
A large variance means that values far from \(\mu\) occur more often.

\bigskip\textsc{Alternative formula for the variance}

It is often convenient to use the formula
\[
\mathrm{Var}(X) = \mathbb{E}[X^2] - \big(\mathbb{E}[X]\big)^2.
\]
Here
\[
\mathbb{E}[X^2] = \int_{-\infty}^{+\infty} x^2 f_X(x) \, dx.
\]
This alternative expression is derived exactly as in the discrete case and is usually easier to use in calculations.

\bigskip\textsc{Effect of linear transformations}

If we transform a random variable \(X\) by a linear function \(aX + b\), the variance behaves as follows:
\[
\mathrm{Var}(aX + b) = a^2 \mathrm{Var}(X).
\]
Adding a constant \(b\) does not change the variance.  
Multiplying by a constant \(a\) scales the variance by \(a^2\).


\subsubsection{Mean and variance of the uniform}

We now illustrate mean and variance calculations using the continuous uniform random variable.  
This is the continuous analogue of the discrete uniform, for which we already know formulas for the mean and variance.

\bigskip\textsc{Mean of the uniform}

Let \(X\) be uniform on \([a,b]\), with density
\[
f_X(x) =
\begin{cases}
\frac{1}{b-a}, & a \le x \le b, \\
0, & \text{otherwise}.
\end{cases}
\]
The mean is
\[
\mathbb{E}[X] = \int_{-\infty}^{+\infty} x f_X(x)\,dx.
\]
Since \(f_X(x)=0\) outside \([a,b]\), this reduces to
\[
\mathbb{E}[X] = \int_a^b x \cdot \frac{1}{b-a}\,dx = \frac{1}{b-a}\int_a^b x\,dx.
\]
We have
\[
\int_a^b x\,dx = \frac{b^2}{2} - \frac{a^2}{2},
\]
so
\[
\mathbb{E}[X] = \frac{1}{b-a}\left(\frac{b^2}{2} - \frac{a^2}{2}\right) = \frac{a+b}{2}.
\]
This is the same as in the discrete case and also matches the midpoint of the interval \([a,b]\), reflecting the symmetry of the PDF around \(\frac{a+b}{2}\).

\bigskip\textsc{Computing \(\mathbb{E}[X^2]\)}

To use the alternative variance formula, we first compute \(\mathbb{E}[X^2]\).  
By the expected value rule,
\[
\mathbb{E}[X^2] = \int_{-\infty}^{+\infty} x^2 f_X(x)\,dx = \int_a^b x^2 \cdot \frac{1}{b-a}\,dx.
\]
Thus
\[
\mathbb{E}[X^2] = \frac{1}{b-a}\int_a^b x^2\,dx.
\]
Since
\[
\int_a^b x^2\,dx = \frac{b^3}{3} - \frac{a^3}{3},
\]
we obtain
\[
\mathbb{E}[X^2] = \frac{1}{b-a}\left(\frac{b^3}{3} - \frac{a^3}{3}\right).
\]
The cubic terms appear because the antiderivative of \(x^2\) is \(\frac{x^3}{3}\).

\bigskip\textsc{Variance and standard deviation}

Using the alternative formula
\[
\mathrm{Var}(X) = \mathbb{E}[X^2] - \big(\mathbb{E}[X]\big)^2,
\]
we substitute
\[
\mathbb{E}[X^2] = \frac{1}{b-a}\left(\frac{b^3}{3} - \frac{a^3}{3}\right),
\quad
\mathbb{E}[X] = \frac{a+b}{2}.
\]
After some algebra, this simplifies to
\[
\mathrm{Var}(X) = \frac{(b-a)^2}{12}.
\]
The standard deviation is then
\[
\sigma = \sqrt{\mathrm{Var}(X)} = \frac{b-a}{\sqrt{12}}.
\]

A useful observation is that the standard deviation is proportional to the width \(b-a\) of the interval.  
The wider the uniform distribution, the larger the standard deviation, which matches the intuition that standard deviation measures the spread or width of the distribution.  
The variance grows with the square of the length \(b-a\), reflecting that doubling the width multiplies the variance by four.


\subsubsection{Exponential random variables}

We now introduce a new continuous random variable, the exponential random variable.  
It is determined by a single positive parameter \(\lambda > 0\), and its PDF is
\[
f_X(x) =
\begin{cases}
\lambda e^{-\lambda x}, & x \ge 0, \\
0, & x < 0.
\end{cases}
\]
Since the PDF is zero for negative \(x\), the exponential random variable is always non-negative.

For \(x \ge 0\), the PDF starts at value \(\lambda\) when \(x = 0\) and then decays exponentially.  
When \(\lambda\) is small, the initial height is small but the decay is slow, so the distribution extends over a wide range.  
When \(\lambda\) is large, the PDF starts high near zero and decays quickly, so most of the probability is concentrated near \(0\).

The shape of the exponential PDF is analogous to that of the geometric distribution in the discrete case.  
There, probabilities decrease geometrically with the value; here they decrease exponentially.

\bigskip\textsc{Tail probabilities}

Fix \(a > 0\).  
We want to compute
\[
\mathbb{P}(X \ge a) = \int_a^{\infty} \lambda e^{-\lambda x} \, dx.
\]
Using the antiderivative \(\int e^{-\lambda x} dx = -\frac{1}{\lambda} e^{-\lambda x}\), we get
\[
\mathbb{P}(X \ge a)
= \lambda \left[-\frac{1}{\lambda} e^{-\lambda x}\right]_{x=a}^{x=\infty}
= - e^{-\lambda x}\Big|_{x=a}^{x=\infty}
= 0 - \big(-e^{-\lambda a}\big)
= e^{-\lambda a}.
\]
Thus the tail probability decays exponentially in \(a\).  

If we set \(a = 0\), we obtain
\[
\mathbb{P}(X \ge 0) = e^{-\lambda \cdot 0} = 1,
\]
which confirms that the total integral of the PDF over its support is \(1\).

\bigskip\textsc{Mean and second moment}

To compute the mean, we use the definition
\[
\mathbb{E}[X] = \int_0^{\infty} x \lambda e^{-\lambda x} \, dx.
\]
This integral can be evaluated by integration by parts and the result is
\[
\mathbb{E}[X] = \frac{1}{\lambda}.
\]

For the second moment, we similarly write
\[
\mathbb{E}[X^2] = \int_0^{\infty} x^2 \lambda e^{-\lambda x} \, dx,
\]
which is again computed by integration by parts (a bit more tedious).  
The result is
\[
\mathbb{E}[X^2] = \frac{2}{\lambda^2}.
\]

\bigskip\textsc{Variance}

Using the alternative variance formula,
\[
\mathrm{Var}(X) = \mathbb{E}[X^2] - \big(\mathbb{E}[X]\big)^2,
\]
we substitute
\[
\mathbb{E}[X^2] = \frac{2}{\lambda^2},
\quad
\mathbb{E}[X] = \frac{1}{\lambda}.
\]
Thus
\[
\mathrm{Var}(X)
= \frac{2}{\lambda^2} - \left(\frac{1}{\lambda}\right)^2
= \frac{2}{\lambda^2} - \frac{1}{\lambda^2}
= \frac{1}{\lambda^2}.
\]

When \(\lambda\) is small, the PDF decays slowly, so large values of \(X\) are more likely.  
In that case, both the mean \(\frac{1}{\lambda}\) and the variance \(\frac{1}{\lambda^2}\) are large, reflecting a wide and spread-out distribution.  

\bigskip\textsc{Interpretation and relation to geometric}

The exponential distribution is closely related to the geometric distribution.  
The mean \(\mathbb{E}[X] = 1/\lambda\) parallels the geometric mean \(1/p\), where \(p\) is the success probability.  
The exponential is often used to model waiting times until an event occurs, such as the time until a customer arrives, a machine fails, a light bulb burns out, or an email arrives.


\subsubsection{Cumulative distribution functions}

We now introduce a way to describe the distribution of any random variable, discrete or continuous, using the \textit{cumulative distribution function} (CDF).  
For a random variable \(X\), the CDF is the function
\[
F_X(x) = \mathbb{P}(X \le x).
\]
It gives the probability that \(X\) takes a value less than or equal to \(x\).

The CDF is defined the same way for all types of random variables.  
If \(X\) is continuous with PDF \(f_X\), then
\[
F_X(x) = \int_{-\infty}^{x} f_X(t)\,dt.
\]

\bigskip\textsc{Example: uniform random variable}

Let \(X\) be uniform on \([a,b]\).  
The CDF \(F_X(x)\) has three cases:
\[
F_X(x) =
\begin{cases}
0, & x \le a, \\
\displaystyle \frac{x-a}{b-a}, & a < x < b, \\
1, & x \ge b.
\end{cases}
\]
For \(x \le a\), there is no probability mass to the left, so the CDF is \(0\).  
For \(a < x < b\), the probability up to \(x\) is the area of a rectangle with base \(x-a\) and height \(\frac{1}{b-a}\), giving \(\frac{x-a}{b-a}\).  
For \(x \ge b\), we have already covered all the probability, so the CDF is \(1\).

The graph starts at \(0\), rises linearly from \(a\) to \(b\), and then stays at \(1\).

\bigskip\textsc{Using the CDF to get probabilities}

Once we know the CDF, we can compute probabilities of intervals.  
For example,
\[
\mathbb{P}(3 < X \le 4) = F_X(4) - F_X(3).
\]
More generally, for any \(a < b\),
\[
\mathbb{P}(a < X \le b) = F_X(b) - F_X(a).
\]
Thus the CDF contains all probabilistic information about the distribution.

For continuous random variables, we can recover the PDF from the CDF wherever the derivative exists:
\[
f_X(x) = \frac{d}{dx} F_X(x).
\]
Conversely, given the PDF, we obtain the CDF by integrating.

\bigskip\textsc{CDFs for discrete random variables}

For a discrete random variable \(X\), the CDF is still
\[
F_X(x) = \mathbb{P}(X \le x),
\]
but it is computed as a sum of PMF values for all outcomes \(k\) with \(k \le x\).  
The CDF has the shape of a staircase: it is constant between the possible values of \(X\) and jumps upward at each value where \(X\) has positive probability.

At a point \(x_0\) where \(X\) can take the value \(x_0\), the jump size is
\[
F_X(x_0) - \lim_{x \uparrow x_0} F_X(x) = \mathbb{P}(X = x_0).
\]
The value of the CDF at a jump is the value \emph{including} that point, because the definition uses \(\le x\).

\bigskip\textsc{General properties of CDFs}

CDFs satisfy some general properties for any random variable:
\[
\text{(i) } F_X(x) \text{ is non-decreasing in } x,
\]
\[
\text{(ii) } \lim_{x \to -\infty} F_X(x) = 0,
\]
\[
\text{(iii) } \lim_{x \to +\infty} F_X(x) = 1.
\]
The non-decreasing property reflects that larger \(x\) correspond to larger events \(\{X \le x\}\).  
As we move far to the left, the probability of being \(\le x\) must go to \(0\), and as we move far to the right, the probability must approach \(1\).

\subsubsection{Normal random variables}

Normal random variables, also called Gaussian random variables, are among the most important in probability theory.  
They play a key role in theoretical results such as the central limit theorem and are widely used in applications as models of random noise.

\bigskip\textsc{Standard normal distribution}

The \textit{standard normal} random variable is denoted by
\[
X \sim \mathcal{N}(0,1).
\]
Its PDF is defined for all real \(x\) by
\[
f_X(x) = \frac{1}{\sqrt{2\pi}} e^{-x^2/2}, \quad -\infty < x < \infty.
\]
This random variable can take any real value.  

The factor \(e^{-x^2/2}\) has a bell shape centered at \(0\): it is equal to \(1\) at \(x=0\) and decreases as \(|x|\) increases.  
The constant \(\frac{1}{\sqrt{2\pi}}\) is chosen so that the total area under the curve is \(1\).

Because the PDF is symmetric around \(0\), the mean of the standard normal is
\[
\mathbb{E}[X] = 0.
\]
A calculus computation shows that its variance is
\[
\mathrm{Var}(X) = 1.
\]

\bigskip\textsc{General normal distributions}

A general normal random variable with mean \(\mu\) and variance \(\sigma^2\) is denoted by
\[
X \sim \mathcal{N}(\mu,\sigma^2),
\]
where \(\sigma > 0\).  
Its PDF is
\[
f_X(x) = \frac{1}{\sqrt{2\pi}\,\sigma}
\exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right), \quad -\infty < x < \infty.
\]

The exponent \(-\frac{(x-\mu)^2}{2\sigma^2}\) is a quadratic centered at \(x=\mu\), so the PDF has a bell shape centered at \(\mu\).  
By symmetry, the mean is
\[
\mathbb{E}[X] = \mu.
\]
A calculus exercise shows that the variance is
\[
\mathrm{Var}(X) = \sigma^2.
\]

The parameter \(\sigma\) controls the spread.  
When \(\sigma\) is small, the quadratic in the exponent rises quickly, the bell is narrow, and the variance is small.  
When \(\sigma\) is large, the bell is wide and the variance is large.

\bigskip\textsc{Linear transformations of normal variables}

Normal random variables behave very nicely under linear transformations.  
Let
\[
X \sim \mathcal{N}(\mu,\sigma^2)
\]
and define
\[
Y = aX + b
\]
for constants \(a\) and \(b\).  

From linearity of expectation,
\[
\mathbb{E}[Y] = a\,\mathbb{E}[X] + b = a\mu + b.
\]
From the variance rule for linear functions,
\[
\mathrm{Var}(Y) = a^2 \mathrm{Var}(X) = a^2 \sigma^2.
\]
The key additional fact is that \(Y\) is itself normal:
\[
Y \sim \mathcal{N}(a\mu + b,\, a^2\sigma^2).
\]

There is one special case.  
If \(a = 0\), then \(Y = b\) is a constant (a degenerate discrete random variable with variance \(0\)).  
By convention, we can view this as a degenerate normal with mean \(b\) and variance \(0\), so that any linear function of a normal random variable is still considered normal.


\subsubsection{Calculation of normal probabilities}

Normal random variables are very important, so we would like to calculate probabilities such as \(\mathbb{P}(X \le 5)\) when \(X\) is normal.  
In general, the CDF of a normal random variable does not have a closed form expression in terms of elementary functions.  

\bigskip\textsc{Standard normal CDF and its table}

For the standard normal random variable \(Y \sim \mathcal{N}(0,1)\), we denote the CDF by
\[
\phi(y) = \mathbb{P}(Y \le y).
\]
Graphically, \(\phi(y)\) is the area under the standard normal PDF to the left of \(y\).  
Values of \(\phi(y)\) are tabulated in standard normal tables.

To use such a table, we locate the row and column corresponding to the first two decimal digits of \(y\).  
For example, \(\phi(0) = 0.5\) by symmetry of the standard normal around \(0\).  
If we want \(\phi(1.16)\), we find the row for \(1.1\) and the column for \(0.06\), and read off the corresponding entry.  
For \(y = 2.9\), we find the row for \(2.9\) and the column for \(0.00\), which gives \(\phi(2.9)\).  

Typical tables list \(\phi(y)\) only for \(y \ge 0\).  
To handle negative arguments, we use symmetry of the standard normal:
\[
\phi(-y) = \mathbb{P}(Y \le -y) = 1 - \mathbb{P}(Y \le y) = 1 - \phi(y), \quad y \ge 0.
\]
Thus, to find \(\phi(-2)\), we compute \(1 - \phi(2)\).  

\bigskip\textsc{Standardization of a general normal}

Let \(X\) be a normal random variable with mean \(\mu\) and standard deviation \(\sigma > 0\), that is \(X \sim \mathcal{N}(\mu,\sigma^2)\).  
We define the standardized random variable
\[
Y = \frac{X - \mu}{\sigma}.
\]
The quantity \(Y\) measures how many standard deviations \(X\) is away from its mean.  

The mean of \(Y\) is
\[
\mathbb{E}[Y] = \frac{\mathbb{E}[X] - \mu}{\sigma} = 0,
\]
and the variance of \(Y\) is
\[
\mathrm{Var}(Y) = \frac{1}{\sigma^2} \mathrm{Var}(X) = 1.
\]
If \(X\) is normal, then \(Y\) is a standard normal random variable.  
Equivalently,
\[
X = \mu + \sigma Y, \quad Y \sim \mathcal{N}(0,1).
\]

\bigskip\textsc{Using standardization to compute probabilities}

To compute probabilities for \(X\), we transform the event into one involving \(Y\).  
For example, for any real \(a\) and \(b\) with \(a < b\),
\[
\mathbb{P}(a \le X \le b)
= \mathbb{P}\!\left(a \le \mu + \sigma Y \le b\right)
= \mathbb{P}\!\left(\frac{a-\mu}{\sigma} \le Y \le \frac{b-\mu}{\sigma}\right).
\]
If we write
\[
\alpha = \frac{a-\mu}{\sigma}, \quad \beta = \frac{b-\mu}{\sigma},
\]
then
\[
\mathbb{P}(a \le X \le b) = \mathbb{P}(\alpha \le Y \le \beta) = \phi(\beta) - \phi(\alpha).
\]
We can then use the standard normal table to find \(\phi(\beta)\) and \(\phi(\alpha)\), handling negative arguments via symmetry if needed.

\bigskip\textsc{Example of the method}

Suppose \(X \sim \mathcal{N}(6,4)\), so \(\mu = 6\) and \(\sigma = 2\).  
We want
\[
\mathbb{P}(2 \le X \le 8).
\]
We standardize:
\[
\frac{2 - 6}{2} = -2, \quad \frac{8 - 6}{2} = 1.
\]
So
\[
\mathbb{P}(2 \le X \le 8) = \mathbb{P}(-2 \le Y \le 1),
\]
where \(Y \sim \mathcal{N}(0,1)\).  
This equals
\[
\mathbb{P}(Y \le 1) - \mathbb{P}(Y \le -2)
= \phi(1) - \phi(-2).
\]
Using symmetry,
\[
\phi(-2) = 1 - \phi(2),
\]
so
\[
\mathbb{P}(2 \le X \le 8)
= \phi(1) - \big(1 - \phi(2)\big)
= \phi(1) + \phi(2) - 1.
\]
Finally, we read \(\phi(1)\) and \(\phi(2)\) from the standard normal table and substitute the corresponding numerical values.


\newpage
\subsubsection{Conditioning a continuous random variable on an event}

We extend conditioning to continuous random variables, by replacing sums with integrals and PMFs with PDFs.

\bigskip\textsc{Conditional PDF on an event}

Let \(X\) be continuous with PDF \(f_X\) and let \(A\) be an event with \(\mathbb{P}(A)>0\).  
In the conditional world where \(A\) is known to have occurred, \(X\) has a conditional PDF \(f_{X\mid A}\) such that, for any set \(B\),
\[
\mathbb{P}(X \in B \mid A) = \int_B f_{X\mid A}(x)\,dx.
\]
This \(f_{X\mid A}\) is a genuine PDF:
\[
f_{X\mid A}(x)\ge 0,\qquad \int_{-\infty}^{\infty} f_{X\mid A}(x)\,dx = 1.
\]

\bigskip\textsc{Special case: conditioning on \(\{X \in S\}\)}

Now assume
\[
A = \{X \in S\}, \quad S \subseteq \mathbb{R}, \quad \mathbb{P}(X \in S)>0.
\]
We want a formula for \(f_{X\mid X\in S}\).

Think in terms of small intervals: for small \(\Delta>0\),
\[
\mathbb{P}(x \le X \le x+\Delta) \approx f_X(x)\,\Delta,
\]
\[
\mathbb{P}(x \le X \le x+\Delta \mid X\in S) \approx f_{X\mid X\in S}(x)\,\Delta.
\]

\textsc{Case 1:} \(x \notin S\).  
For small \(\Delta\), the interval \([x,x+\Delta]\) does not intersect \(S\), so the conditional probability is \(0\), hence
\[
f_{X\mid X\in S}(x)=0 \quad \text{if } x\notin S.
\]

\textsc{Case 2:} \(x \in S\).  
For small \(\Delta\), \([x,x+\Delta]\subseteq S\), so
\[
\mathbb{P}(x \le X \le x+\Delta \mid X\in S)
=
\frac{\mathbb{P}(x \le X \le x+\Delta)}{\mathbb{P}(X\in S)}
\approx
\frac{f_X(x)\,\Delta}{\mathbb{P}(X\in S)}.
\]
Comparing with
\[
\mathbb{P}(x \le X \le x+\Delta \mid X\in S) \approx f_{X\mid X\in S}(x)\,\Delta
\]
and cancelling \(\Delta\), we get
\[
f_{X\mid X\in S}(x) = \frac{f_X(x)}{\mathbb{P}(X\in S)} \quad \text{if } x\in S,
\]
and \(0\) otherwise.

In compact form,
\[
f_{X\mid X\in S}(x)
=
\frac{f_X(x)}{\int_S f_X(u)\,du}\,\mathbf{1}_{\{x\in S\}}.
\]
\textit{The conditional PDF keeps the same shape as \(f_X\) on \(S\), but is rescaled so that the total area over \(S\) becomes \(1\).}

\bigskip\textsc{Conditional expectation (brief)}

Given \(A\) with \(\mathbb{P}(A)>0\), the conditional expectation of \(g(X)\) given \(A\) is
\[
\mathbb{E}[g(X)\mid A] = \int_{-\infty}^{\infty} g(x)\,f_{X\mid A}(x)\,dx.
\]
For the special case \(A=\{X\in S\}\),
\[
\mathbb{E}[g(X)\mid X\in S]
=
\frac{1}{\mathbb{P}(X\in S)}\int_S g(x)\,f_X(x)\,dx.
\]

This is the continuous analogue of the discrete conditional expectation formula, with integrals instead of sums.


\subsubsection{Conditioning example}

We illustrate conditioning of a continuous random variable on an event using a simple piecewise constant PDF.

\bigskip\textsc{Setup}

Let \(X\) have a uniform PDF on \([a,b]\),
\[
f_X(x) = \frac{1}{b-a}, \quad a \le x \le b,
\]
and \(f_X(x)=0\) otherwise.  
We condition on the event
\[
A = \left\{ \frac{a+b}{2} \le X \le b \right\}.
\]

\bigskip\textsc{Conditional PDF}

The conditional PDF is
\[
f_{X\mid A}(x) = 0 \quad \text{for } x < \frac{a+b}{2} \text{ or } x > b.
\]
On the interval \(\left[\frac{a+b}{2}, b\right]\), it must keep the same \textit{shape} as \(f_X\), which is constant, so \(f_{X\mid A}\) is also constant there.  
The length of the conditioning interval is
\[
b - \frac{a+b}{2} = \frac{b-a}{2}.
\]
Since the total area must be \(1\),
\[
\text{height} \times \text{length} = 1
\quad\Rightarrow\quad
\text{height} \cdot \frac{b-a}{2} = 1,
\]
so the conditional PDF height is
\[
\text{height} = \frac{2}{b-a}.
\]
Therefore
\[
f_{X\mid A}(x)
=
\begin{cases}
\dfrac{2}{b-a}, & \dfrac{a+b}{2} \le x \le b,\\[4pt]
0, & \text{otherwise}.
\end{cases}
\]

\bigskip\textsc{Conditional expectation}

Because \(X\mid A\) is uniform on \(\left[\frac{a+b}{2},b\right]\), its expectation is the midpoint of that interval:
\[
\mathbb{E}[X \mid A]
=
\frac{1}{2}\left(\frac{a+b}{2}\right)
+
\frac{1}{2}b
=
\frac{1}{4}a + \frac{3}{4}b.
\]

\bigskip\textsc{Conditional expectation of \(X^2\)}

Using the expected value rule with the conditional PDF,
\[
\mathbb{E}[X^2 \mid A]
=
\int_{-\infty}^{\infty} x^2 f_{X\mid A}(x)\,dx
=
\int_{(a+b)/2}^{b} x^2 \cdot \frac{2}{b-a}\,dx.
\]
That is,
\[
\mathbb{E}[X^2 \mid A]
=
\frac{2}{b-a}
\int_{(a+b)/2}^{b} x^2\,dx,
\]
an elementary integral that can be evaluated directly if needed.

\subsubsection{Exponential PDF and memorylessness}

Let \(T\) be the lifetime of a light bulb and assume \(T \sim \text{Exponential}(\lambda)\), so
\[
\mathbb{P}(T > x) = e^{-\lambda x}, \quad x \ge 0.
\]

\bigskip\textsc{Memorylessness}

Suppose the bulb has already worked for time \(t\) and has not failed, so \(T>t\).  
Let \(X = T - t\) be the remaining lifetime. Then
\[
\mathbb{P}(X > x \mid T > t)
=
\frac{\mathbb{P}(T > t + x)}{\mathbb{P}(T > t)}
=
\frac{e^{-\lambda (t+x)}}{e^{-\lambda t}}
=
e^{-\lambda x}
=
\mathbb{P}(T > x).
\]
So the remaining lifetime has the same distribution as a new bulb.  
\textit{An exponential lifetime does not “remember” how long it has already lasted.}

\bigskip\textsc{Small-interval interpretation}

For small \(\Delta>0\),
\[
\mathbb{P}(0 \le T \le \Delta) \approx \lambda \Delta.
\]
By memorylessness, the same is true conditionally at any time \(t\):
\[
\mathbb{P}(\text{failure in }[t,t+\Delta) \mid T>t) \approx \lambda \Delta.
\]
\textit{At each tiny time step, the failure chance is always about \(\lambda \Delta\), independent of the past.}


\subsubsection{Total probability and expectation theorems (continuous case)}

We now state continuous versions of the total probability and total expectation theorems, and illustrate them with a simple scenario model.

\bigskip\textsc{Total probability for PDFs}

Let \(A_1,\dots,A_n\) be a partition of the sample space, that is, disjoint events whose union is the whole space, and let \(X\) be a continuous random variable with density \(f_X\).  
For each scenario \(A_i\), let \(f_{X\mid A_i}\) be the conditional PDF of \(X\) given \(A_i\).

Then the PDF of \(X\) is
\[
f_X(x) = \sum_{i=1}^n \mathbb{P}(A_i)\,f_{X\mid A_i}(x).
\]
\textit{The overall density at a point \(x\) is a weighted sum of the conditional densities at \(x\), with weights given by the scenario probabilities.}

\bigskip\textsc{Total expectation (continuous)}

Under the same setup, the total expectation theorem says
\[
\mathbb{E}[X]
=
\sum_{i=1}^n \mathbb{P}(A_i)\,\mathbb{E}[X \mid A_i].
\]
\textit{The overall mean is the average of the conditional means, weighted by the probabilities of the scenarios.}

\bigskip\textsc{Example: Bill and the supermarket}

Bill will go to the supermarket at a random time \(X\) (in hours from now).  
There are two scenarios:

\textsc{Scenario 1} \(A_1\): with probability \(\mathbb{P}(A_1)=\tfrac{1}{3}\), Bill goes between 0 and 2 hours, uniformly:
\[
f_{X\mid A_1}(x) =
\begin{cases}
\dfrac{1}{2}, & 0 \le x \le 2,\\
0, & \text{otherwise}.
\end{cases}
\]

\textsc{Scenario 2} \(A_2\): with probability \(\mathbb{P}(A_2)=\tfrac{2}{3}\), Bill goes between 6 and 8 hours, uniformly:
\[
f_{X\mid A_2}(x) =
\begin{cases}
\dfrac{1}{2}, & 6 \le x \le 8,\\
0, & \text{otherwise}.
\end{cases}
\]

Using the total probability formula,
\[
f_X(x)
=
\mathbb{P}(A_1)f_{X\mid A_1}(x)
+
\mathbb{P}(A_2)f_{X\mid A_2}(x),
\]
so
\[
f_X(x)
=
\begin{cases}
\dfrac{1}{3}\cdot \dfrac{1}{2} = \dfrac{1}{6}, & 0 \le x \le 2,\\[4pt]
\dfrac{2}{3}\cdot \dfrac{1}{2} = \dfrac{1}{3}, & 6 \le x \le 8,\\[4pt]
0, & \text{otherwise}.
\end{cases}
\]

For the expectation, use total expectation:
\[
\mathbb{E}[X]
=
\mathbb{P}(A_1)\,\mathbb{E}[X\mid A_1]
+
\mathbb{P}(A_2)\,\mathbb{E}[X\mid A_2].
\]
Under \(A_1\), \(X\) is uniform on \([0,2]\), so \(\mathbb{E}[X\mid A_1]=1\).  
Under \(A_2\), \(X\) is uniform on \([6,8]\), so \(\mathbb{E}[X\mid A_2]=7\).  
Thus
\[
\mathbb{E}[X]
=
\frac{1}{3}\cdot 1 + \frac{2}{3}\cdot 7
=
\frac{1}{3} + \frac{14}{3}
=
5.
\]

This example shows how scenario-wise conditional PDFs and expectations combine to give the overall distribution and mean.


\subsubsection{Mixed random variables}

We now consider random variables that are neither purely discrete nor purely continuous.

\bigskip\textsc{Motivating example}

You have 1 dollar and can choose whether to play a lottery.

- With probability \(1/2\), you do nothing and keep your 1 dollar.  
  In this scenario, \(X=1\) for sure.  
- With probability \(1/2\), you play the lottery and get a random return \(U\) that is uniform on \([0,2]\).  
  In this scenario, \(X=U\).

The resulting \(X\) can take any value in \([0,2]\), but
\[
\mathbb{P}(X = 1) = \frac{1}{2} > 0.
\]
So:

- \(X\) is \textit{not} discrete, because it is not supported on a countable set only  
- \(X\) is \textit{not} continuous, because a continuous random variable must satisfy \(\mathbb{P}(X=x)=0\) for all \(x\), which fails at \(x=1\)

We call such an \(X\) a \textit{mixed random variable}.

\bigskip\textsc{General construction}

More generally, suppose:

- With probability \(p\), \(X\) behaves like a discrete random variable \(Y\)  
- With probability \(1-p\), \(X\) behaves like a continuous random variable \(Z\)

Then \(X\) has no PMF (not purely discrete) and no PDF (not purely continuous), but its CDF is always well defined.  
Let \(F_Y\) and \(F_Z\) be the CDFs of \(Y\) and \(Z\).  
Using the total probability theorem for CDFs,
\[
F_X(x)
=
\mathbb{P}(X \le x)
=
p\,\mathbb{P}(Y \le x) + (1-p)\,\mathbb{P}(Z \le x)
=
p\,F_Y(x) + (1-p)\,F_Z(x).
\]
\textit{The CDF of a mixed random variable is a weighted sum of the CDFs of its discrete and continuous parts.}

For the expectation, we define
\[
\mathbb{E}[X]
=
p\,\mathbb{E}[Y] + (1-p)\,\mathbb{E}[Z],
\]
consistent with the total expectation theorem.

\bigskip\textsc{CDF in the lottery example}

In the lottery example:

- Scenario \(A_1\) (play): probability \(1/2\), \(X\) is uniform on \([0,2]\)  
  \[
  f_{X\mid A_1}(x) =
  \begin{cases}
  \dfrac{1}{2}, & 0 \le x \le 2,\\
  0, & \text{otherwise},
  \end{cases}
  \]
  so
  \[
  F_{X\mid A_1}(x) =
  \begin{cases}
  0, & x < 0,\\
  \dfrac{x}{2}, & 0 \le x \le 2,\\
  1, & x > 2.
  \end{cases}
  \]

- Scenario \(A_2\) (do nothing): probability \(1/2\), \(X\equiv 1\)  
  Then
  \[
  F_{X\mid A_2}(x) =
  \begin{cases}
  0, & x < 1,\\
  1, & x \ge 1.
  \end{cases}
  \]

Using
\[
F_X(x)
=
\frac{1}{2}F_{X\mid A_1}(x) + \frac{1}{2}F_{X\mid A_2}(x),
\]
we get:

- For \(x<0\): \(F_X(x)=0\)  
- For \(0 \le x < 1\):
  \[
  F_X(x) = \frac{1}{2}\cdot \frac{x}{2} + \frac{1}{2}\cdot 0 = \frac{x}{4},
  \]
  a line of slope \(1/4\)  
- At \(x=1\): we add the jump from the discrete part
  \[
  \lim_{x\to 1^-}F_X(x) = \frac{1}{4}, \qquad F_X(1) = \frac{1}{2}\cdot\frac{1}{2} + \frac{1}{2}\cdot 1 = \frac{1}{4} + \frac{1}{2} = \frac{3}{4},
  \]
  so the jump size at \(x=1\) is \(3/4 - 1/4 = 1/2\), which equals \(\mathbb{P}(X=1)\)  
- For \(1 < x \le 2\):
  \[
  F_X(x) = \frac{1}{2}\cdot\frac{x}{2} + \frac{1}{2}\cdot 1 = \frac{x}{4} + \frac{1}{2},
  \]
  again a line of slope \(1/4\) until it reaches 1 at \(x=2\)  
- For \(x>2\): \(F_X(x)=1\)

The resulting CDF has:

- Linear growth with slope \(1/4\) on \([0,1)\)  
- A jump of size \(1/2\) at \(x=1\)  
- Linear growth with slope \(1/4\) on \((1,2]\), then flat

\textit{This illustrates that a mixed random variable may not have a single PDF or PMF, but its CDF is easy to construct from the component distributions.}


\subsubsection{Joint PDFs}

We now move to \textit{multiple} continuous random variables and introduce the joint PDF, the continuous analogue of the joint PMF.

\bigskip\textsc{Definition and probability of a set}

Let \(X\) and \(Y\) be random variables.  
A \textit{joint PDF} is a function \(f_{X,Y}(x,y)\) such that

\[
f_{X,Y}(x,y) \ge 0 \quad \text{for all } (x,y),
\]
\[
\iint_{\mathbb{R}^2} f_{X,Y}(x,y)\,dx\,dy = 1.
\]

If \(X\) and \(Y\) are \textit{jointly continuous}, probabilities are computed by
\[
\mathbb{P}\big((X,Y)\in B\big)
=
\iint_B f_{X,Y}(x,y)\,dx\,dy,
\]
for any region \(B\subseteq\mathbb{R}^2\).  
\textit{Geometrically, this is the volume under the surface \(z=f_{X,Y}(x,y)\) above the set \(B\).}

\bigskip\textsc{Rectangles and small rectangles}

For a rectangle
\[
B = \{(x,y)\colon a \le x \le b,\ c \le y \le d\},
\]
we write
\[
\mathbb{P}(a \le X \le b,\ c \le Y \le d)
=
\int_c^{d}\int_a^{b} f_{X,Y}(x,y)\,dx\,dy.
\]

If the rectangle is very small, say \([x, x+\Delta]\times[y, y+\Delta]\) with small \(\Delta>0\), then
\[
\mathbb{P}\big(x \le X \le x+\Delta,\ y \le Y \le y+\Delta\big)
\approx
f_{X,Y}(x,y)\,\Delta^2.
\]
So \textit{the joint PDF is a probability density per unit area}: probability \(\approx\) density \(\times\) area.

\bigskip\textsc{Zero-area sets}

For jointly continuous random variables, any set of area \(0\) has probability \(0\).  
In particular:

- \(\mathbb{P}(X = x_0,\ Y = y_0) = 0\) for any point \((x_0,y_0)\)  
- \(\mathbb{P}\big( (X,Y)\) lies on a specific curve or line\()\big) = 0\)

\bigskip\textsc{When continuity fails jointly}

Consider \(X\) continuous and define \(Y=X\).  
Each variable alone is continuous, but the pair \((X,Y)\) always lies on the line \(y=x\).  
All probability is concentrated on this one-dimensional line, which has area \(0\).  

This contradicts the \textit{jointly continuous} picture where zero-area sets must have probability \(0\).  
So \(X\) and \(Y\) here are \textit{not} jointly continuous, even though each is continuous by itself.


\newpage
\subsubsection{From the joint to the marginal}

We now see how to recover the individual (marginal) PDFs of \(X\) and \(Y\) from their joint PDF.

\bigskip\textsc{Deriving the marginal of \(X\)}

Let \(X,Y\) be jointly continuous with joint PDF \(f_{X,Y}(x,y)\).  
The CDF of \(X\) is
\[
F_X(x) = \mathbb{P}(X \le x).
\]
Geometrically, this is the probability that the point \((X,Y)\) falls in the half-plane \(\{(s,t)\colon s \le x\}\).  
Using the joint PDF,
\[
F_X(x)
=
\iint_{\{(s,t)\colon s \le x\}} f_{X,Y}(s,t)\,ds\,dt.
\]

We can write this as an iterated integral (use \(s\) for the \(X\)-coordinate and \(t\) for the \(Y\)-coordinate):
\[
F_X(x)
=
\int_{-\infty}^{\infty}
\left(
\int_{-\infty}^{x} f_{X,Y}(s,t)\,ds
\right)dt.
\]

To get the PDF of \(X\), we differentiate:
\[
f_X(x)
=
\frac{d}{dx}F_X(x)
=
\frac{d}{dx}
\int_{-\infty}^{\infty}
\left(
\int_{-\infty}^{x} f_{X,Y}(s,t)\,ds
\right)dt.
\]
Differentiating w.r.t. the upper limit \(x\) of the inner integral gives
\[
f_X(x)
=
\int_{-\infty}^{\infty} f_{X,Y}(x,t)\,dt.
\]

So the \textit{marginal PDF of \(X\)} is obtained by integrating the joint PDF over all possible \(y\)-values:
\[
f_X(x)
=
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dy.
\]
Similarly,
\[
f_Y(y)
=
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dx.
\]

\bigskip\textsc{Example: uniform joint PDF on a region}

Suppose \((X,Y)\) has a joint PDF that is uniform on a set \(S\) in the plane, and zero outside.  
If \(\text{area}(S)=A\), then
\[
f_{X,Y}(x,y)
=
\begin{cases}
\dfrac{1}{A}, & (x,y)\in S,\\[4pt]
0, & \text{otherwise}.
\end{cases}
\]

For any set \(B\),
\[
\mathbb{P}\big((X,Y)\in B\big)
=
\iint_{B} f_{X,Y}(x,y)\,dx\,dy
=
\frac{1}{A}\,\text{area}(B\cap S).
\]

Now fix some \(x\).  
To find \(f_X(x)\), integrate along the vertical line at that \(x\):
\[
f_X(x)
=
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dy
=
\frac{1}{A}\times \big(\text{length of the segment of this line inside }S\big).
\]
If that vertical slice inside \(S\) has length \(L(x)\), then
\[
f_X(x) = \frac{L(x)}{A}.
\]

Similarly, for \(f_Y(y)\), we integrate horizontally; if the horizontal slice at \(y\) has length \(M(y)\) inside \(S\), then
\[
f_Y(y) = \frac{M(y)}{A}.
\]

\textit{In words: the marginal at a given point is the joint density times the “thickness” of \(S\) along the corresponding line, i.e., the length of the slice through \(S\).}


\subsubsection{Continuous analogs of various properties}

We now briefly summarize how standard discrete results extend to jointly continuous random variables.

\bigskip\textsc{Joint PDFs for many variables}

For more than two continuous random variables \(X_1,\dots,X_n\), we can define a joint PDF
\[
f_{X_1,\dots,X_n}(x_1,\dots,x_n),
\]
which is nonnegative and integrates to 1 over \(\mathbb{R}^n\).  
Probabilities of events in \(\mathbb{R}^n\) are obtained by integrating this joint PDF over the corresponding region.

\bigskip\textsc{From joint to marginals}

The marginal PDF of any subset of the variables is obtained by integrating out the others.  
For example, for the first variable,
\[
f_{X_1}(x_1) = \int_{-\infty}^{\infty}\cdots\int_{-\infty}^{\infty}
f_{X_1,\dots,X_n}(x_1,\dots,x_n)\,dx_2\cdots dx_n.
\]

\bigskip\textsc{Functions of jointly continuous variables}

If \(X_1,\dots,X_n\) are jointly continuous and \(g\) is a real function, we can define
\[
Y = g(X_1,\dots,X_n).
\]
Then the expected value rule has the same form as in the discrete case:
\[
\mathbb{E}[Y]
=
\mathbb{E}\big[g(X_1,\dots,X_n)\big]
=
\int\cdots\int g(x_1,\dots,x_n)\,
f_{X_1,\dots,X_n}(x_1,\dots,x_n)\,dx_1\cdots dx_n.
\]

\bigskip\textsc{Linearity of expectation}

From this rule we recover the usual linearity properties.  
For constants \(a,b\) and random variables \(X,Y\),
\[
\mathbb{E}[aX + bY] = a\,\mathbb{E}[X] + b\,\mathbb{E}[Y],
\]
with no extra assumptions needed.  

\textit{In summary, all familiar discrete laws carry over to the continuous case by the usual recipe: replace sums by integrals and PMFs by PDFs.}


\subsubsection{Joint CDFs}

We can also describe the distribution of multiple random variables using a joint CDF, in complete analogy with the one-dimensional case.

\bigskip\textsc{Definition and relation to the joint PDF}

For a pair \((X,Y)\), the \textit{joint CDF} is
\[
F_{X,Y}(x,y)
=
\mathbb{P}(X \le x,\ Y \le y).
\]
Geometrically, this is the probability that \((X,Y)\) falls in the rectangle \((-\infty,x]\times(-\infty,y]\).

If \(X,Y\) are jointly continuous with joint PDF \(f_{X,Y}\), then
\[
F_{X,Y}(x,y)
=
\int_{-\infty}^{y}
\int_{-\infty}^{x}
f_{X,Y}(s,t)\,ds\,dt.
\]

Conversely, we can recover the joint PDF by differentiating:
\[
f_{X,Y}(x,y)
=
\frac{\partial^2}{\partial x\,\partial y} F_{X,Y}(x,y),
\]
assuming the derivatives exist.

\bigskip\textsc{Example: uniform on the unit square}

Suppose \((X,Y)\) is uniform on the unit square \([0,1]\times[0,1]\).  
Then
\[
f_{X,Y}(x,y)
=
\begin{cases}
1, & 0 \le x \le 1,\ 0 \le y \le 1,\\
0, & \text{otherwise}.
\end{cases}
\]

For \(0 \le x \le 1,\ 0 \le y \le 1\), the joint CDF is the probability of the rectangle \([0,x]\times[0,y]\), which is just its area:
\[
F_{X,Y}(x,y) = xy.
\]

Differentiating,
\[
\frac{\partial^2}{\partial x\,\partial y}F_{X,Y}(x,y)
=
\frac{\partial}{\partial y}\big(y\big)
=
1,
\]
which recovers the joint PDF on the unit square, as expected.




\subsubsection{Conditional PDFs of one variable given another}

Let \(X\) and \(Y\) be jointly continuous with joint PDF \(f_{X,Y}(x,y)\) and marginal \(f_Y(y)\).  
For \(y\) such that \(f_Y(y) > 0\), the \textit{conditional PDF} of \(X\) given \(Y=y\) is
\[
f_{X\mid Y}(x\mid y)
=
\frac{f_{X,Y}(x,y)}{f_Y(y)}.
\]

\textsc{Small-interval interpretation}

For small \(\delta > 0\), we interpret
\[
\mathbb{P}(x \le X \le x + \delta \mid Y \approx y)
\approx
f_{X\mid Y}(x\mid y)\,\delta.
\]
So, given that \(Y\) is (approximately) \(y\), the function \(x \mapsto f_{X\mid Y}(x\mid y)\) plays the same role as an ordinary PDF of \(X\).

\textsc{Conditional probabilities of sets}

For any set \(B \subseteq \mathbb{R}\), we define
\[
\mathbb{P}(X \in B \mid Y = y)
\triangleq
\int_B f_{X\mid Y}(x\mid y)\,dx,
\]
and treat this as the conditional probability of \(X\) lying in \(B\), given that \(Y\) took value \(y\).

\textsc{Example}

Suppose
\[
f_{X,Y}(x,y)
=
\begin{cases}
2, & 0 \le x \le y \le 1,\\
0, & \text{otherwise}.
\end{cases}
\]
Then for \(0 \le y \le 1\),
\[
f_Y(y)
=
\int_0^y 2\,dx
=
2y,
\]
and, for \(0<y\le 1\),
\[
f_{X\mid Y}(x\mid y)
=
\frac{f_{X,Y}(x,y)}{f_Y(y)}
=
\frac{2}{2y}
=
\frac{1}{y},
\qquad 0 \le x \le y,
\]
so conditionally on \(Y=y\), the random variable \(X\) is uniform on \([0,y]\).


\subsubsection{Comments on conditional PDFs}

\textsc{Basic properties}

For jointly continuous \(X,Y\) with joint PDF \(f_{X,Y}(x,y)\) and marginal \(f_Y(y)\), recall
\[
f_{X\mid Y}(x\mid y)
=
\frac{f_{X,Y}(x,y)}{f_Y(y)}
\qquad\text{for } f_Y(y)>0.
\]
This conditional PDF is always nonnegative, because the joint PDF is nonnegative and the denominator is positive.

\textsc{Slices of the joint PDF}

Fix a value \(y\) with \(f_Y(y)>0\).  
As a function of \(x\), the numerator \(f_{X,Y}(x,y)\) describes how the joint density varies along the horizontal line \(\{(x,y)\colon x\in\mathbb{R}\}\).  
Thus \(x \mapsto f_{X,Y}(x,y)\) is a \textit{slice} of the joint surface taken at height \(y\).

The conditional PDF \(f_{X\mid Y}(x\mid y)\) is obtained from this slice by multiplying by the constant factor \(1/f_Y(y)\).  
So, for fixed \(y\), the conditional PDF has the \textit{same shape} in \(x\) as the corresponding slice of the joint PDF, just rescaled vertically.

\textsc{Why it integrates to 1}

For any fixed \(y\) with \(f_Y(y)>0\),
\[
\int_{-\infty}^{\infty} f_{X\mid Y}(x\mid y)\,dx
=
\int_{-\infty}^{\infty} \frac{f_{X,Y}(x,y)}{f_Y(y)}\,dx
=
\frac{1}{f_Y(y)}
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dx.
\]
By the marginalization rule,
\[
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dx
=
f_Y(y),
\]
so the ratio is equal to 1.  
This shows that, for each fixed \(y\), the conditional PDF behaves like an ordinary PDF in the variable \(x\): it is nonnegative and integrates to 1.

\textsc{Multiplication rule for densities}

Starting again from
\[
f_{X\mid Y}(x\mid y)
=
\frac{f_{X,Y}(x,y)}{f_Y(y)},
\]
we can rearrange to obtain
\[
f_{X,Y}(x,y)
=
f_{X\mid Y}(x\mid y)\,f_Y(y).
\]
This has the same form as the multiplication rule for probabilities, where we write
\(\mathbb{P}(A\cap B) = \mathbb{P}(A\mid B)\,\mathbb{P}(B)\).  
The only difference is that here we are dealing with densities, not probabilities, but the algebraic structure is identical.

By symmetry, we also have
\[
f_{X,Y}(x,y)
=
f_{Y\mid X}(y\mid x)\,f_X(x).
\]
So, at the level of formulas, conditional PDFs mirror all the familiar rules from the discrete case, with PMFs replaced by PDFs.



\subsubsection{Total probability and total expectation theorems}

\textsc{Continuous total probability}

For jointly continuous \(X,Y\) and any set \(A \subseteq \mathbb{R}\),
\[
\mathbb{P}(X \in A)
=
\int_{-\infty}^{\infty}
\mathbb{P}(X \in A \mid Y = y)\,f_Y(y)\,dy.
\]

\textsc{Conditional expectation}

For \(f_Y(y)>0\), the conditional mean of \(X\) given \(Y=y\) is
\[
\mathbb{E}[X \mid Y = y]
=
\int_{-\infty}^{\infty} x\,f_{X\mid Y}(x\mid y)\,dx.
\]

\textsc{Total expectation}

The continuous total expectation theorem (law of total expectation) is
\[
\mathbb{E}[X]
=
\int_{-\infty}^{\infty}
\mathbb{E}[X \mid Y = y]\,f_Y(y)\,dy.
\]

\textsc{Expected value rule in the conditional universe}

For any function \(g\),
\[
\mathbb{E}\big[g(X) \mid Y = y\big]
=
\int_{-\infty}^{\infty} g(x)\,f_{X\mid Y}(x\mid y)\,dx.
\]


\subsubsection{Independence}

\textsc{Definition}

Let \(X\) and \(Y\) be jointly continuous with joint PDF \(f_{X,Y}(x,y)\) and marginals \(f_X(x)\), \(f_Y(y)\).  
They are \textit{independent} if
\[
f_{X,Y}(x,y) = f_X(x)\,f_Y(y)
\quad\text{for all }x,y.
\]

\textsc{Relation to conditional PDFs}

Whenever \(f_Y(y)>0\), the multiplication rule gives
\[
f_{X,Y}(x,y) = f_{X\mid Y}(x\mid y)\,f_Y(y).
\]
If \(X\) and \(Y\) are independent, comparing with
\[
f_{X,Y}(x,y) = f_X(x)\,f_Y(y)
\]
yields
\[
f_{X\mid Y}(x\mid y) = f_X(x)
\quad\text{for all }x\text{ and all }y\text{ with }f_Y(y)>0.
\]
So independence means that conditioning on \(Y=y\) does not change the distribution of \(X\).

\textsc{Slice picture}

For fixed \(y\), the conditional PDF \(f_{X\mid Y}(x\mid y)\) is (up to scaling) a slice of the joint PDF along the horizontal line at height \(y\).  
Independence requires that all these slices, as functions of \(x\), have exactly the same shape, namely the shape of the marginal PDF \(f_X\).

\textsc{Intuition}

Under independence, no matter what value of \(Y\) we observe, the distribution of \(X\) remains the same.  
In this sense, \(Y\) carries no information about \(X\), and by symmetry, \(X\) carries no information about \(Y\).

\textsc{More than two variables}

Random variables \(X_1,\dots,X_n\) are independent if their joint PDF factors as
\[
f_{X_1,\dots,X_n}(x_1,\dots,x_n)
=
\prod_{i=1}^n f_{X_i}(x_i)
\quad\text{for all }(x_1,\dots,x_n).
\]
Intuitively, knowing some of the values \(X_i\) does not affect our beliefs about the others.

\textsc{Consequences}

If \(X\) and \(Y\) are independent, then
\[
\mathbb{E}[XY] = \mathbb{E}[X]\,\mathbb{E}[Y],
\]
and, for a sum,
\[
\operatorname{Var}(X+Y) = \operatorname{Var}(X) + \operatorname{Var}(Y).
\]
More generally, if \(X\) and \(Y\) are independent and \(g\) and \(h\) are measurable functions, then \(g(X)\) and \(h(Y)\) are independent, and
\[
\mathbb{E}\big[g(X)\,h(Y)\big]
=
\mathbb{E}[g(X)]\,\mathbb{E}[h(Y)].
\]

\textsc{Example}

Suppose \(X\) and \(Y\) are independent, each uniform on \([0,1]\).  
Then
\[
f_X(x) =
\begin{cases}
1, & 0 \le x \le 1,\\
0, & \text{otherwise},
\end{cases}
\qquad
f_Y(y) =
\begin{cases}
1, & 0 \le y \le 1,\\
0, & \text{otherwise},
\end{cases}
\]
and
\[
f_{X,Y}(x,y) = f_X(x)\,f_Y(y)
=
\begin{cases}
1, & 0 \le x \le 1,\ 0 \le y \le 1,\\
0, & \text{otherwise}.
\end{cases}
\]
Here every horizontal slice in \(x\) has the same flat shape, matching the marginal density of \(X\), which reflects the independence of \(X\) and \(Y\).


\subsubsection{Stick-breaking example}

\textsc{Setup}

A stick has length \(l\).  
First break: the location \(X\) is uniform on \([0,l]\), so
\[
f_X(x)
=
\begin{cases}
\frac{1}{l}, & 0 \le x \le l,\\
0, & \text{otherwise}.
\end{cases}
\]
Second break: given \(X=x\), the location \(Y\) is uniform on \([0,x]\), so
\[
f_{Y\mid X}(y\mid x)
=
\begin{cases}
\frac{1}{x}, & 0 \le y \le x,\\
0, & \text{otherwise}.
\end{cases}
\]

\textsc{Dependence}

The conditional PDF of \(Y\) changes when \(x\) changes (its support is \([0,x]\) and its height is \(1/x\)), so \(Y\) clearly depends on \(X\).  
Equivalently, knowing \(X=x\) tells us that \(0 \le Y \le x\), so \(X\) provides information about \(Y\), hence \(X\) and \(Y\) are not independent.

\textsc{Joint and marginal PDFs}

Using the multiplication rule,
\[
f_{X,Y}(x,y)
=
f_X(x)\,f_{Y\mid X}(y\mid x)
=
\frac{1}{l}\cdot\frac{1}{x}
=
\frac{1}{lx},
\]
for the \textit{possible} pairs \((x,y)\), namely
\[
0 \le y \le x \le l,
\]
and \(f_{X,Y}(x,y)=0\) otherwise.  
This corresponds to a joint density supported on the triangle \(\{(x,y)\colon 0\le y\le x\le l\}\).

To find the marginal of \(Y\), fix \(y\) and integrate over the admissible \(x\):
\[
f_Y(y)
=
\int_{-\infty}^{\infty} f_{X,Y}(x,y)\,dx
=
\int_{y}^{l} \frac{1}{lx}\,dx
=
\frac{1}{l}\big[\log x\big]_{x=y}^{x=l}
=
\frac{1}{l}\log\!\Big(\frac{l}{y}\Big),
\]
for \(0<y\le l\), and \(f_Y(y)=0\) otherwise.  
This density blows up as \(y\to 0\), but it is still valid because its total area is 1.

\textsc{Expected value of \(Y\)}

A direct computation uses
\[
\mathbb{E}[Y]
=
\int_{0}^{l} y\,f_Y(y)\,dy
=
\int_{0}^{l} y\cdot\frac{1}{l}\log\!\Big(\frac{l}{y}\Big)\,dy,
\]
but this integral is messy.  
A simpler route is to use the total expectation theorem and condition on \(X\):
\[
\mathbb{E}[Y]
=
\int_{0}^{l} \mathbb{E}[Y\mid X=x]\,f_X(x)\,dx.
\]
Given \(X=x\), the variable \(Y\) is uniform on \([0,x]\), so
\[
\mathbb{E}[Y\mid X=x] = \frac{x}{2}.
\]
Thus
\[
\mathbb{E}[Y]
=
\int_{0}^{l} \frac{x}{2}\cdot\frac{1}{l}\,dx
=
\frac{1}{2l}\int_{0}^{l} x\,dx
=
\frac{1}{2l}\cdot\frac{l^{2}}{2}
=
\frac{l}{4}.
\]
\textit{Intuitively, each break cuts the expected length by a factor of \(1/2\), so after two breaks the expected remaining length is \(l/4\).}


\subsubsection{Independent normals}

\textsc{Standard independent normals}

Let \(X\) and \(Y\) be independent standard normal random variables.  
Their joint PDF is the product of the marginals:
\[
f_{X,Y}(x,y)
=
\frac{1}{\sqrt{2\pi}}e^{-x^{2}/2}
\cdot
\frac{1}{\sqrt{2\pi}}e^{-y^{2}/2}
=
\frac{1}{2\pi}\exp\!\Big(-\frac{1}{2}(x^{2}+y^{2})\Big).
\]
For any fixed radius \(r\), the circle \(\{(x,y)\colon x^{2}+y^{2}=r^{2}\}\) is a contour on which the joint PDF is constant, so the contours are circles centered at the origin.

\textsc{General independent normals}

Now let \(X\sim\mathcal{N}(\mu_{x},\sigma_{x}^{2})\) and \(Y\sim\mathcal{N}(\mu_{y},\sigma_{y}^{2})\) be independent.  
Their joint PDF is
\[
f_{X,Y}(x,y)
=
\frac{1}{2\pi\sigma_{x}\sigma_{y}}
\exp\!\Bigg(
-\frac{1}{2}\Bigg[
\frac{(x-\mu_{x})^{2}}{\sigma_{x}^{2}}
+
\frac{(y-\mu_{y})^{2}}{\sigma_{y}^{2}}
\Bigg]
\Bigg).
\]
The prefactor \(1/(2\pi\sigma_{x}\sigma_{y})\) is just the normalizing constant.  
Contours correspond to level sets of the quadratic form
\[
\frac{(x-\mu_{x})^{2}}{\sigma_{x}^{2}}
+
\frac{(y-\mu_{y})^{2}}{\sigma_{y}^{2}}
=
\text{constant},
\]
which are ellipses centered at \((\mu_{x},\mu_{y})\) with principal axes aligned with the coordinate axes.

\textsc{Shape and stretching}

The joint PDF attains its maximum at \((x,y)=(\mu_{x},\mu_{y})\), where the exponent is zero.  
Moving to ellipses farther from this center decreases the value of the joint PDF, but it remains constant along each ellipse.

If \(\sigma_{y}^{2} > \sigma_{x}^{2}\), then the ellipses are more stretched in the \(y\)-direction than in the \(x\)-direction, reflecting that \(Y\) typically varies over a wider range than \(X\).  
In the special case \(\sigma_{x}^{2} = \sigma_{y}^{2}\), these ellipses become circles (after appropriate rescaling), recovering a circularly symmetric bell.

\textsc{Independence versus dependence}

For independent normals, the stretching of the bell is always along the coordinate axes.  
If the contours were ellipses tilted away from the axes (with principal axes along some diagonal direction), this would indicate dependence between \(X\) and \(Y\), leading to a general bivariate normal distribution rather than a product of independent one-dimensional normals.

\subsubsection{Bayes rule variations}

\textsc{Discrete case}

In the discrete setting we have a hidden random variable \(X\) with a known prior PMF \(p_X(x)\), and an observable random variable \(Y\) related to \(X\) through the conditional PMF \(p_{Y\mid X}(y\mid x)\).  
After observing \(Y=y\), Bayes rule gives the posterior PMF of \(X\):
\[
p_{X\mid Y}(x\mid y)
=
\frac{p_{Y\mid X}(y\mid x)\,p_X(x)}{p_Y(y)},
\]
where the marginal PMF of \(Y\) is
\[
p_Y(y)
=
\sum_{x} p_{Y\mid X}(y\mid x)\,p_X(x).
\]

\textsc{Continuous case}

In the continuous setting the same structure carries over with PMFs replaced by PDFs and sums by integrals.  
If \(X\) has prior PDF \(f_X(x)\) and \(Y\) has conditional PDF \(f_{Y\mid X}(y\mid x)\), then after observing \(Y=y\) the posterior PDF of \(X\) is
\[
f_{X\mid Y}(x\mid y)
=
\frac{f_{Y\mid X}(y\mid x)\,f_X(x)}{f_Y(y)},
\]
with
\[
f_Y(y)
=
\int_{-\infty}^{\infty} f_{Y\mid X}(y\mid x)\,f_X(x)\,dx.
\]

\textsc{Interpretation}

In both cases, Bayes rule updates prior beliefs about \(X\) using the likelihood of the observation \(Y=y\).  
The numerator combines prior and likelihood into a joint model, while the denominator normalizes over all possible values of \(X\) via the total probability theorem, ensuring the posterior is a valid distribution.


\subsubsection{Detection of a binary signal}

\textsc{Model}

A binary random variable \(K\) takes values \(\pm 1\) with equal prior probabilities
\[
\mathbb{P}(K=1)=\mathbb{P}(K=-1)=\frac{1}{2}.
\]
The observation is
\[
Y = K + W,
\]
where \(W\) is standard normal, independent of \(K\).  
Given \(K=k\), the random variable \(Y\) is normal with mean \(k\) and variance \(1\), so
\[
f_{Y\mid K}(y\mid k)
=
\frac{1}{\sqrt{2\pi}}
\exp\!\Big(-\frac{1}{2}(y-k)^{2}\Big),
\qquad k\in\{-1,1\}.
\]

\textsc{Posterior probability via Bayes rule}

We want the posterior probability that \(K=1\) given an observed value \(Y=y\):
\[
\mathbb{P}(K=1\mid Y=y)
=
\frac{f_{Y\mid K}(y\mid 1)\,\mathbb{P}(K=1)}
{f_{Y\mid K}(y\mid 1)\,\mathbb{P}(K=1)
+
f_{Y\mid K}(y\mid -1)\,\mathbb{P}(K=-1)}.
\]
Using the equal priors, this becomes
\[
\mathbb{P}(K=1\mid Y=y)
=
\frac{f_{Y\mid K}(y\mid 1)}
{f_{Y\mid K}(y\mid 1)+f_{Y\mid K}(y\mid -1)}.
\]
Substituting the Gaussian PDFs and simplifying the exponentials yields
\[
\mathbb{P}(K=1\mid Y=y)
=
\frac{1}{1+\exp(-2y)}.
\]

\textsc{Interpretation}

As a function of \(y\), the posterior \(\mathbb{P}(K=1\mid Y=y)\) is a smooth S-shaped curve:
\[
\mathbb{P}(K=1\mid Y=y)\to 1 \ \text{as }y\to+\infty,\qquad
\mathbb{P}(K=1\mid Y=y)\to 0 \ \text{as }y\to-\infty,
\]
and
\[
\mathbb{P}(K=1\mid Y=0)=\frac{1}{2}.
\]
Large positive \(y\) makes it very likely that \(K=1\), large negative \(y\) makes it very likely that \(K=-1\), and \(y=0\) leaves both possibilities equally likely.


\subsubsection{Inference of the bias of a coin}

\textsc{Model}

We observe a Bernoulli random variable \(K\in\{0,1\}\) with
\[
\mathbb{P}(K=1\mid Y=y)=y,
\qquad
\mathbb{P}(K=0\mid Y=y)=1-y,
\]
where the (unknown) bias \(Y\) itself is random and modeled as uniform on \([0,1]\), so
\[
f_Y(y)
=
\begin{cases}
1, & 0 \le y \le 1,\\
0, & \text{otherwise}.
\end{cases}
\]

\textsc{Goal}

Given a single observation \(K=1\), we want the posterior density \(f_{Y\mid K}(y\mid 1)\).

\textsc{Bayes rule}

Using Bayes rule with a continuous \(Y\) and discrete \(K\),
\[
f_{Y\mid K}(y\mid 1)
=
\frac{\mathbb{P}(K=1\mid Y=y)\,f_Y(y)}
{\mathbb{P}(K=1)}.
\]
Here
\[
\mathbb{P}(K=1\mid Y=y)=y,
\qquad
f_Y(y)=1 \ \text{on }[0,1],
\]
and
\[
\mathbb{P}(K=1)
=
\int_{0}^{1} \mathbb{P}(K=1\mid Y=u)\,f_Y(u)\,du
=
\int_{0}^{1} u\,du
=
\frac{1}{2}.
\]
Therefore, for \(0\le y\le 1\),
\[
f_{Y\mid K}(y\mid 1)
=
\frac{y\cdot 1}{1/2}
=
2y,
\]
and \(f_{Y\mid K}(y\mid 1)=0\) otherwise.

\textsc{Interpretation}

Initially, \(Y\) was uniform, so all biases in \([0,1]\) were equally plausible.  
After observing \(K=1\), the posterior density \(f_{Y\mid K}(y\mid 1)=2y\) tilts toward larger \(y\), reflecting that higher biases make the outcome \(K=1\) more likely.